---
title: "Stroke Prediction"
author: "Anurag Nepal"
date: "2025-03-16"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
library(tidyverse)
library(ggplot2)
```

# Background

I wanted to see how well stroke occurrence can be predicted from patient demographics and health indicators, and which modeling choices matter for a rare event. The plan: compare eight classifiers across four preprocessing scenarios, with and without SMOTE for the class imbalance and with and without PCA, and judge each on AUC, accuracy, sensitivity, and specificity.

# Setup

The data is `healthcare-dataset-stroke-data.csv`, in this repo. One row per patient: age, gender, hypertension, heart disease, marital and work status, residence type, glucose level, BMI, smoking status, and the stroke indicator. Only about 5% of rows are strokes, so the imbalance drives most of the modeling decisions below.

## Exploratory analysis

```{r}
stroke_data = read.csv("healthcare-dataset-stroke-data.csv")
```

```{r}
glimpse(stroke_data)
summary(stroke_data)
```

BMI should be integer but is there as character.
```{r}
stroke_data$bmi=as.double(stroke_data$bmi)
summary(stroke_data)
```
```{r}
stroke_counts = table(stroke_data$stroke)
stroke_counts_df = as.data.frame(stroke_counts)
stroke_counts_df$percentage = stroke_counts_df$Freq / sum(stroke_counts_df$Freq) * 100
colnames(stroke_counts_df) = c("stroke", "count", "percentage")
print(stroke_counts_df)
```
Data imbalance so may have to augment the data.

```{r}
na_counts = colSums(is.na(stroke_data))
na_counts[na_counts > 0]
```
```{r}
stroke_data = stroke_data[stroke_data$gender != "Other", ]
```


```{r}
categorize_age = function(age) {
  if (age >= 0 & age <= 1) {
    return('Infant')
  } else if (age > 1 & age <= 12) {
    return('Children')
  } else if (age > 12 & age < 20) {
    return('Teenager')
  } else if (age >= 20 & age <= 39) {
    return('Adult')
  } else if (age > 39 & age <= 59) {
    return('Middle Aged')
  } else {
    return('Senior')
  }
}

stroke_data$age_group = sapply(stroke_data$age, categorize_age)
unique(stroke_data$age_group)
```

```{r}
avg_bmi_by_group = aggregate(bmi ~ age_group + gender, data = stroke_data, FUN = mean)
colnames(avg_bmi_by_group) = c("age_group","gender","avg_bmi")
avg_bmi_by_group
```
```{r}
# Keep the pre-imputation values. The group means above are computed on the
# full dataset, so this fill is for the EDA plots only. The modeling section
# redoes the imputation leakage-free (means from the training split only).
stroke_data$bmi_raw = stroke_data$bmi
stroke_data = merge(stroke_data, avg_bmi_by_group, by = c("age_group", "gender"), all.x = TRUE)
stroke_data$bmi = ifelse(is.na(stroke_data$bmi), stroke_data$avg_bmi, stroke_data$bmi)
stroke_data$avg_bmi = NULL
```
Used the average bmi for the age group and gender to fill in the bmi NA values for the exploratory plots below. Before modeling, the imputation is redone with group means from the training split only, so no test information leaks into training.

```{r}
boxplot(age ~ stroke, data = stroke_data, main = "Age by Stroke Status", col = "skyblue")
boxplot(avg_glucose_level ~ stroke, data = stroke_data, main = "Glucose Level by Stroke Status", col = "lightgreen")
boxplot(bmi ~ stroke, data = stroke_data, main = "BMI by Stroke Status", col = "salmon")
```


```{r}
barplot(sort(table(stroke_data$age_group)), beside = TRUE, main = "Distribution by Age Group",
        xlab = "Age Group", ylab = "Count", col = c("skyblue", "salmon"),
        legend = rownames(stroke_data$age_group))
age_group_stroke = prop.table((table(stroke_data$age_group, stroke_data$stroke)), margin = 1)
barplot(sort(age_group_stroke[, 2]), col = "skyblue", main = "Stroke Rate by Age Group", ylab = "Stroke Proportion")
```
```{r}
hist(stroke_data$age, main = "Distribution of Age", xlab = "Age", col = "skyblue", breaks = 20, prob = TRUE)
lines(density(stroke_data$age, na.rm = TRUE), col = "black", lwd = 3)
legend("topleft", legend = c("Histogram", "Density"), col = c("skyblue", "black"), lwd = 2)

plot(density(stroke_data$age[stroke_data$stroke == 1]), col = "red", main = "Age Distribution by Stroke Status")
lines(density(stroke_data$age[stroke_data$stroke == 0]), col = "blue")
legend("topleft", legend = c("No Stroke","Stroke"), fill = c("blue", "red"))
```
```{r}
hist(stroke_data$avg_glucose_level, main = "Distribution of Glucose Level", xlab = "Glucose Level", col = "lightgreen", breaks = 20,prob = TRUE)
lines(density(stroke_data$avg_glucose_level), col = "black",lwd=3)
legend("topright", legend = c("Histogram", "Density"), col = c("lightgreen", "black"), lwd = 2)
plot(density(stroke_data$avg_glucose_level[stroke_data$stroke == 0]), col = "blue", main = "Glucose Level Distribution by Stroke Status")
lines(density(stroke_data$avg_glucose_level[stroke_data$stroke == 1]), col = "red")
legend("topright", legend = c("No Stroke","Stroke"), fill = c("blue", "red"))
```
```{r}
hist(stroke_data$bmi, main = "Distribution of BMI", xlab = "BMI", col = "salmon", breaks = 20,prob=TRUE)
lines(density(stroke_data$bmi), col = "black", main = "BMI Distribution",lwd=3)
legend("topright", legend = c("Histogram", "Density"), col = c("salmon", "black"), lwd = 2)
plot(density(stroke_data$bmi[stroke_data$stroke == 1]), col = "red", main = "BMI Distribution by Stroke Status")
lines(density(stroke_data$bmi[stroke_data$stroke == 0]), col = "blue")
legend("topright", legend = c("No Stroke","Stroke"), fill = c("blue", "red"))
```

Stroke likelihood increases with lower glucose concentration.
```{r}
barplot(table(stroke_data$gender), main = "Gender Distribution", col = "skyblue")
gender_stroke = prop.table(table(stroke_data$gender, stroke_data$stroke), margin = 1)
barplot(gender_stroke[, 2], col = "skyblue", main = "Stroke Rate by Gender", ylab = "Stroke Proportion")
```

```{r}
barplot(table(stroke_data$smoking_status), main = "Smoking Status Distribution", col = "orange")
smoke_stroke = prop.table(table(stroke_data$smoking_status, stroke_data$stroke), margin = 1)
barplot(smoke_stroke[, 2], col = "skyblue", main = "Stroke Rate by Smoking Status", ylab = "Stroke Proportion")
```
Former smokers have higher stroke rates.
```{r}
barplot(table(stroke_data$heart_disease),names.arg = c("No Heart Disease", "Heart Disease"), main = "Heart Disease Distribution", col = "salmon")
hd_stroke = prop.table(table(stroke_data$heart_disease, stroke_data$stroke), margin = 1)
barplot(hd_stroke[, 2], names.arg = c("No Heart Disease", "Heart Disease"), col = "pink", main = "Stroke Rate by Heart Disease")
```
Heart disease increases stroke risk.


```{r}
stroke_data %>%
  group_by(work_type) %>%
  summarise(hypertension_rate = mean(hypertension)) %>%
  ggplot(aes(work_type, hypertension_rate)) + 
  geom_col(fill = "orange") + 
  labs(title = "Hypertension Rate by Work Type")
```

```{r}
stroke_data %>%
  group_by(ever_married) %>%
  summarise(hypertension_rate = mean(hypertension)) %>%
  ggplot(aes(ever_married, hypertension_rate)) + 
  geom_col(fill = "lightblue") + 
  labs(title = "Hypertension Rate by Marriage Status")
```
```{r}
stroke_data %>%
  group_by(Residence_type) %>%
  summarise(hypertension_rate = mean(hypertension)) %>%
  ggplot(aes(Residence_type, hypertension_rate)) + 
  geom_col(fill = "lightblue") + 
  labs(title = "Hypertension Rate by Residence Type")
```

```{r}
barplot(table(stroke_data$hypertension),names.arg = c("No Hypertension", "Hypertension"), main = "Hypertension Distribution", col = "lightgreen")
hypertension_stroke = prop.table(table(stroke_data$hypertension, stroke_data$stroke), margin = 1)
barplot(hypertension_stroke[, 2], names.arg = c("No Hypertension", "Hypertension"), col = "skyblue", main = "Stroke Rate by Hypertension")
```
```{r}
barplot(table(stroke_data$work_type), main = "Work Type Distribution", col = "orange")
res_stroke = prop.table(table(stroke_data$work_type, stroke_data$stroke), margin = 1)
barplot(res_stroke[, 2], col = "salmon", main = "Stroke Rate by Work Type")
```
```{r}
barplot(table(stroke_data$Residence_type), main = "Residence Distribution", col = "orange")
res_stroke = prop.table(table(stroke_data$Residence_type, stroke_data$stroke), margin = 1)
barplot(res_stroke[, 2], names.arg = c("Rural", "Urban"), col = "salmon", main = "Stroke Rate by Residence")
```
```{r}
barplot(table(stroke_data$ever_married),names.arg = c("Never Married", "Married"), main = " Marital Status Distribution", col = "lightblue")
married_stroke = prop.table(table(stroke_data$ever_married, stroke_data$stroke), margin = 1)
barplot(married_stroke[, 2], names.arg = c("Never Married", "Married"), col = "purple", main = "Stroke Rate by Marital Status")
```

```{r}
numerical_vars = stroke_data[, c("age", "avg_glucose_level", "bmi")]
numerical_vars$stroke = stroke_data$stroke
cor_matrix = cor(numerical_vars, use = "complete.obs")
print(cor_matrix)
```



# Method

Eight models, four scenarios each: logistic regression, LDA, QDA, KNN, decision tree, random forest, bagging, and boosting. Every model trains under 5-fold cross-validation optimizing ROC, then gets scored on a held-out 30% test split. The four scenarios cross SMOTE (yes/no) with PCA (yes/no).

## Training controls

```{r}
library(caret)
library(pROC)
```

```{r}
# Stash the pre-imputation BMI and grouping columns for the leakage-free
# imputation below: they are dropped from the modeling data in the next chunk.
bmi_lookup = stroke_data[, c("age_group", "gender", "bmi_raw")]
stroke_data = stroke_data %>%
  select(-id, -age_group,-Residence_type,-gender, -bmi_raw) %>%
  mutate(
    ever_married = as.factor(ever_married),
    work_type = as.factor(work_type),
    smoking_status = as.factor(smoking_status),
    hypertension = as.factor(hypertension),
    heart_disease = as.factor(heart_disease),
    stroke = as.factor(stroke)
  )
stroke_data = stroke_data %>%
  mutate(stroke = factor(stroke, levels = c(1, 0), labels = c("YES", "NO")))
```


```{r}
set.seed(123)  
trainIndex = createDataPartition(stroke_data$stroke, p = 0.7, list = FALSE)
train_data = stroke_data[trainIndex, ]
test_data = stroke_data[-trainIndex, ]
prop.table(table(train_data$stroke))
prop.table(table(test_data$stroke))
```

```{r leakfree-bmi}
# Leakage-free BMI imputation. The EDA fill used group means computed on the
# full dataset, which lets test information leak into the training rows. Redo
# it here: group means come from the training split only, then overwrite bmi
# in both splits. Row order matches trainIndex, so the lookup aligns exactly.
train_lookup = bmi_lookup[trainIndex, ]
test_lookup = bmi_lookup[-trainIndex, ]
train_bmi_means = aggregate(bmi_raw ~ age_group + gender, data = train_lookup,
                            FUN = function(x) mean(x, na.rm = TRUE))
colnames(train_bmi_means) = c("age_group", "gender", "avg_bmi")
overall_bmi = mean(train_lookup$bmi_raw, na.rm = TRUE)
train_bmi_means$avg_bmi[is.nan(train_bmi_means$avg_bmi)] = overall_bmi
fill_bmi = function(lookup) {
  key_l = paste(lookup$age_group, lookup$gender)
  key_m = paste(train_bmi_means$age_group, train_bmi_means$gender)
  fill = train_bmi_means$avg_bmi[match(key_l, key_m)]
  fill[is.na(fill)] = overall_bmi
  ifelse(is.na(lookup$bmi_raw), fill, lookup$bmi_raw)
}
train_data$bmi = fill_bmi(train_lookup)
test_data$bmi = fill_bmi(test_lookup)
stopifnot(sum(is.na(train_data$bmi)) == 0, sum(is.na(test_data$bmi)) == 0)
```

Two control objects carry the design. We'll define two `trainControl` objects for 5-fold cross-validation: one without SMOTE and one with SMOTE for class balancing. We'll use these across all models and scenarios.

```{r train-controls}
# Without SMOTE
ctrl_base = trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,           
  summaryFunction = twoClassSummary,  
)

ctrl_smote = trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  sampling = "smote"           
)
```


For each model (Logistic Regression, LDA, QDA, KNN), we'll train under four scenarios:
1. **No SMOTE, No PCA**
2. **SMOTE, No PCA**
3. **No SMOTE, PCA**
4. **SMOTE, PCA**


## The models

I start with the linear classifiers.

### Logistic regression

```{r}
# No SMOTE, No PCA
set.seed(123)
logit_no_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "glm", 
  family = "binomial",
  trControl = ctrl_base, 
  metric = "ROC"
)
logit_pred_no_smote_no_pca_prob = predict(logit_no_smote_no_pca, newdata = test_data, type = "prob")
logit_pred_no_smote_no_pca_class = predict(logit_no_smote_no_pca, newdata = test_data, type = "raw")
roc_logit_no_smote_no_pca = roc(test_data$stroke, logit_pred_no_smote_no_pca_prob[,"YES"])
acc_logit_no_smote_no_pca = confusionMatrix(logit_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]


# SMOTE, No PCA
set.seed(123)
logit_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "glm", 
  family = "binomial",
  trControl = ctrl_smote, 
  metric = "ROC"
)
logit_pred_smote_no_pca_prob = predict(logit_smote_no_pca, newdata = test_data, type = "prob")
logit_pred_smote_no_pca_class = predict(logit_smote_no_pca, newdata = test_data, type = "raw")
roc_logit_smote_no_pca = roc(test_data$stroke, logit_pred_smote_no_pca_prob[,"YES"])
acc_logit_smote_no_pca = confusionMatrix(logit_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]


# No SMOTE, PCA
set.seed(123)
logit_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "glm", 
  family = "binomial",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),  # PCA with centering and scaling
  metric = "ROC"
)
logit_pred_no_smote_pca_prob = predict(logit_no_smote_pca, newdata = test_data, type = "prob")
logit_pred_no_smote_pca_class = predict(logit_no_smote_pca, newdata = test_data, type = "raw")
roc_logit_no_smote_pca = roc(test_data$stroke, logit_pred_no_smote_pca_prob[,"YES"])
acc_logit_no_smote_pca = confusionMatrix(logit_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, PCA
set.seed(123)
logit_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "glm", 
  family = "binomial",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
logit_pred_smote_pca_prob = predict(logit_smote_pca, newdata = test_data, type = "prob")
roc_logit_smote_pca = roc(test_data$stroke, logit_pred_smote_pca_prob[,"YES"])
logit_pred_smote_pca_class = predict(logit_smote_pca, newdata = test_data, type = "raw")
acc_logit_smote_pca = confusionMatrix(logit_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# Plot ROC curves
plot(roc_logit_no_smote_no_pca, col = "blue", main = "ROC for Logistic Regression")
lines(roc_logit_smote_no_pca, col = "red")
lines(roc_logit_no_smote_pca, col = "green")
lines(roc_logit_smote_pca, col = "purple")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 
       col = c("blue", "red", "green", "purple"), 
       lty = 1)
```

### Linear discriminant analysis

```{r}
# No SMOTE, No PCA
set.seed(123)
lda_no_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "lda",
  trControl = ctrl_base, 
  metric = "ROC"
)
lda_pred_no_smote_no_pca = predict(lda_no_smote_no_pca, newdata = test_data, type = "prob")
roc_lda_no_smote_no_pca = roc(test_data$stroke, lda_pred_no_smote_no_pca[, "YES"])
lda_pred_no_smote_no_pca_class = predict(lda_no_smote_no_pca, newdata = test_data, type = "raw")
acc_lda_no_smote_no_pca = confusionMatrix(lda_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]


# SMOTE, No PCA
set.seed(123)
lda_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "lda",
  trControl = ctrl_smote, 
  metric = "ROC"
)
lda_pred_smote_no_pca = predict(lda_smote_no_pca, newdata = test_data, type = "prob")
roc_lda_smote_no_pca = roc(test_data$stroke, lda_pred_smote_no_pca[, "YES"])
lda_pred_smote_no_pca_class = predict(lda_smote_no_pca, newdata = test_data, type = "raw")
acc_lda_smote_no_pca = confusionMatrix(lda_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]


# No SMOTE, PCA
set.seed(123)
lda_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "lda",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
lda_pred_no_smote_pca = predict(lda_no_smote_pca, newdata = test_data, type = "prob")
roc_lda_no_smote_pca = roc(test_data$stroke, lda_pred_no_smote_pca[, "YES"])
lda_pred_no_smote_pca_class = predict(lda_no_smote_pca, newdata = test_data, type = "raw")
acc_lda_no_smote_pca = confusionMatrix(lda_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]


# SMOTE, PCA
set.seed(123)
lda_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "lda",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
lda_pred_smote_pca = predict(lda_smote_pca, newdata = test_data, type = "prob")
roc_lda_smote_pca = roc(test_data$stroke, lda_pred_smote_pca[, "YES"])
lda_pred_smote_pca_class = predict(lda_smote_pca, newdata = test_data, type = "raw")
acc_lda_smote_pca = confusionMatrix(lda_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]
# Plot ROC curves
plot(roc_lda_no_smote_no_pca, col = "blue", main = "ROC for LDA")
lines(roc_lda_smote_no_pca, col = "red")
lines(roc_lda_no_smote_pca, col = "green")
lines(roc_lda_smote_pca, col = "purple")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 
       col = c("blue", "red", "green", "purple"), 
       lty = 1)
```

### Quadratic discriminant analysis

```{r}

# No SMOTE, No PCA
set.seed(123)
qda_no_smote_no_pca = tryCatch({
  train(
    stroke ~ ., 
    data = train_data, 
    method = "qda",
    trControl = ctrl_base, 
    metric = "ROC"
  )
}, error = function(e) NULL)
qda_pred_no_smote_no_pca = if (!is.null(qda_no_smote_no_pca)) {
  predict(qda_no_smote_no_pca, newdata = test_data, type = "prob")
} else {
  NULL
}
roc_qda_no_smote_no_pca = if (!is.null(qda_pred_no_smote_no_pca)) {
  roc(test_data$stroke, qda_pred_no_smote_no_pca[, "YES"])
} else {
  NULL
}
qda_pred_no_smote_no_pca_class = if (!is.null(qda_no_smote_no_pca)) {
  predict(qda_no_smote_no_pca, newdata = test_data, type = "raw")
} else {
  NULL
}
acc_qda_no_smote_no_pca = if (!is.null(qda_pred_no_smote_no_pca_class)) {
  confusionMatrix(qda_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]
} else {
  NA
}

# SMOTE, No PCA
set.seed(123)
qda_smote_no_pca = tryCatch({
  train(
    stroke ~ ., 
    data = train_data, 
    method = "qda",
    trControl = ctrl_smote, 
    metric = "ROC"
  )
}, error = function(e) NULL)
qda_pred_smote_no_pca = if (!is.null(qda_smote_no_pca)) {
  predict(qda_smote_no_pca, newdata = test_data, type = "prob")
} else {
  NULL
}
roc_qda_smote_no_pca = if (!is.null(qda_pred_smote_no_pca)) {
  roc(test_data$stroke, qda_pred_smote_no_pca[, "YES"])
} else {
  NULL
}
qda_pred_smote_no_pca_class = if (!is.null(qda_smote_no_pca)) {
  predict(qda_smote_no_pca, newdata = test_data, type = "raw")
} else {
  NULL
}
acc_qda_smote_no_pca = if (!is.null(qda_pred_smote_no_pca_class)) {
  confusionMatrix(qda_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]
} else {
  NA
}

# No SMOTE, PCA
set.seed(123)
qda_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "qda",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
qda_pred_no_smote_pca = predict(qda_no_smote_pca, newdata = test_data, type = "prob")
roc_qda_no_smote_pca = roc(test_data$stroke, qda_pred_no_smote_pca[, "YES"])
qda_pred_no_smote_pca_class = predict(qda_no_smote_pca, newdata = test_data, type = "raw")
acc_qda_no_smote_pca = confusionMatrix(qda_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, PCA
set.seed(123)
qda_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "qda",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
qda_pred_smote_pca = predict(qda_smote_pca, newdata = test_data, type = "prob")
roc_qda_smote_pca = roc(test_data$stroke, qda_pred_smote_pca[, "YES"])
qda_pred_smote_pca_class = predict(qda_smote_pca, newdata = test_data, type = "raw")
acc_qda_smote_pca = confusionMatrix(qda_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# Plot ROC curves (only for successful models)
plot(roc_qda_no_smote_pca, col = "green", main = "ROC for QDA")
lines(roc_qda_smote_pca, col = "purple")
if (!is.null(roc_qda_no_smote_no_pca)) lines(roc_qda_no_smote_no_pca, col = "blue")
if (!is.null(roc_qda_smote_no_pca)) lines(roc_qda_smote_no_pca, col = "red")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA")[c(!is.null(roc_qda_no_smote_no_pca), !is.null(roc_qda_smote_no_pca), TRUE, TRUE)], 
       col = c("blue", "red", "green", "purple")[c(!is.null(roc_qda_no_smote_no_pca), !is.null(roc_qda_smote_no_pca), TRUE, TRUE)], 
       lty = 1)

```


### K-nearest neighbors

```{r}

# No SMOTE, No PCA
set.seed(123)
knn_no_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "knn",
  trControl = ctrl_base, 
  metric = "ROC"
)
knn_pred_no_smote_no_pca = predict(knn_no_smote_no_pca, newdata = test_data, type = "prob")
roc_knn_no_smote_no_pca = roc(test_data$stroke, knn_pred_no_smote_no_pca[, "YES"])
knn_pred_no_smote_no_pca_class = predict(knn_no_smote_no_pca, newdata = test_data, type = "raw")
acc_knn_no_smote_no_pca = confusionMatrix(knn_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, No PCA
set.seed(123)
knn_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "knn",
  trControl = ctrl_smote, 
  metric = "ROC"
)
knn_pred_smote_no_pca = predict(knn_smote_no_pca, newdata = test_data, type = "prob")
roc_knn_smote_no_pca = roc(test_data$stroke, knn_pred_smote_no_pca[, "YES"])
knn_pred_smote_no_pca_class = predict(knn_smote_no_pca, newdata = test_data, type = "raw")
acc_knn_smote_no_pca = confusionMatrix(knn_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# No SMOTE, PCA
set.seed(123)
knn_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "knn",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
knn_pred_no_smote_pca = predict(knn_no_smote_pca, newdata = test_data, type = "prob")
roc_knn_no_smote_pca = roc(test_data$stroke, knn_pred_no_smote_pca[, "YES"])
knn_pred_no_smote_pca_class = predict(knn_no_smote_pca, newdata = test_data, type = "raw")
acc_knn_no_smote_pca = confusionMatrix(knn_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, PCA
set.seed(123)
knn_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "knn",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
knn_pred_smote_pca = predict(knn_smote_pca, newdata = test_data, type = "prob")
roc_knn_smote_pca = roc(test_data$stroke, knn_pred_smote_pca[, "YES"])
knn_pred_smote_pca_class = predict(knn_smote_pca, newdata = test_data, type = "raw")
acc_knn_smote_pca = confusionMatrix(knn_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# Plot ROC curves
plot(roc_knn_no_smote_no_pca, col = "blue", main = "ROC for KNN")
lines(roc_knn_smote_no_pca, col = "red")
lines(roc_knn_no_smote_pca, col = "green")
lines(roc_knn_smote_pca, col = "purple")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 
       col = c("blue", "red", "green", "purple"), 
       lty = 1)
```

```{r}
library(gbm)        # For boosting (gbm)
library(rpart)      # For decision trees
library(ipred)      # For bagging
library(tibble)     # For AUC data frame
library(dplyr)      # For data manipulation
library(knitr)      # For table formatting
```

The remaining four are tree-based: a single decision tree, random forest, bagging, and boosting.

### Decision tree

```{r}
#Decision Tree
# No SMOTE, No PCA
set.seed(123)
dt_no_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rpart",
  trControl = ctrl_base, 
  tuneLength = 5,  # Tune complexity parameter
  metric = "ROC"
)
dt_pred_no_smote_no_pca_prob = predict(dt_no_smote_no_pca, newdata = test_data, type = "prob")
dt_pred_no_smote_no_pca_class = predict(dt_no_smote_no_pca, newdata = test_data, type = "raw")
roc_dt_no_smote_no_pca = roc(test_data$stroke, dt_pred_no_smote_no_pca_prob[,"YES"])
acc_dt_no_smote_no_pca = confusionMatrix(dt_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]


# SMOTE, No PCA
set.seed(123)
dt_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rpart",
  trControl = ctrl_smote, 
  tuneLength = 5,
  metric = "ROC"
)
dt_pred_smote_no_pca_prob = predict(dt_smote_no_pca, newdata = test_data, type = "prob")
dt_pred_smote_no_pca_class = predict(dt_smote_no_pca, newdata = test_data, type = "raw")
roc_dt_smote_no_pca = roc(test_data$stroke, dt_pred_smote_no_pca_prob[,"YES"])
acc_dt_smote_no_pca = confusionMatrix(dt_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]


# No SMOTE, PCA
set.seed(123)
dt_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rpart",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),
  tuneLength = 5,
  metric = "ROC"
)
dt_pred_no_smote_pca_prob = predict(dt_no_smote_pca, newdata = test_data, type = "prob")
dt_pred_no_smote_pca_class = predict(dt_no_smote_pca, newdata = test_data, type = "raw")
roc_dt_no_smote_pca = roc(test_data$stroke, dt_pred_no_smote_pca_prob[,"YES"])
acc_dt_no_smote_pca = confusionMatrix(dt_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]


# SMOTE, PCA
set.seed(123)
dt_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rpart",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  tuneLength = 5,
  metric = "ROC"
)
dt_pred_smote_pca_prob = predict(dt_smote_pca, newdata = test_data, type = "prob")
dt_pred_smote_pca_class = predict(dt_smote_pca, newdata = test_data, type = "raw")
roc_dt_smote_pca = roc(test_data$stroke, dt_pred_smote_pca_prob[,"YES"])
acc_dt_smote_pca = confusionMatrix(dt_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]


# Plot ROC curves
plot(roc_dt_no_smote_no_pca, col = "blue", main = "ROC for Decision Tree")
lines(roc_dt_smote_no_pca, col = "red")
lines(roc_dt_no_smote_pca, col = "green")
lines(roc_dt_smote_pca, col = "purple")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 
       col = c("blue", "red", "green", "purple"), 
       lty = 1)
```

### Random forest

```{r}
#Random Forest
# No SMOTE, No PCA
set.seed(123)
rf_no_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rf",
  trControl = ctrl_base, 
  tuneLength = 3,  # Tune mtry
  metric = "ROC"
)
rf_pred_no_smote_no_pca = predict(rf_no_smote_no_pca, newdata = test_data, type = "prob")
roc_rf_no_smote_no_pca = roc(test_data$stroke, rf_pred_no_smote_no_pca[,"YES"])
rf_pred_no_smote_no_pca_class = predict(rf_no_smote_no_pca, newdata = test_data, type = "raw")
acc_rf_no_smote_no_pca = confusionMatrix(rf_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, No PCA
set.seed(123)
rf_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rf",
  trControl = ctrl_smote, 
  tuneLength = 3,
  metric = "ROC"
)
rf_pred_smote_no_pca = predict(rf_smote_no_pca, newdata = test_data, type = "prob")
roc_rf_smote_no_pca = roc(test_data$stroke, rf_pred_smote_no_pca[,"YES"])
rf_pred_smote_no_pca_class = predict(rf_smote_no_pca, newdata = test_data, type = "raw")
acc_rf_smote_no_pca = confusionMatrix(rf_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# No SMOTE, PCA
set.seed(123)
rf_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rf",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),
  tuneLength = 3,
  metric = "ROC"
)
rf_pred_no_smote_pca = predict(rf_no_smote_pca, newdata = test_data, type = "prob")
roc_rf_no_smote_pca = roc(test_data$stroke, rf_pred_no_smote_pca[,"YES"])
rf_pred_no_smote_pca_class = predict(rf_no_smote_pca, newdata = test_data, type = "raw")
acc_rf_no_smote_pca = confusionMatrix(rf_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, PCA
set.seed(123)
rf_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "rf",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  tuneLength = 3,
  metric = "ROC"
)
rf_pred_smote_pca = predict(rf_smote_pca, newdata = test_data, type = "prob")
roc_rf_smote_pca = roc(test_data$stroke, rf_pred_smote_pca[,"YES"])
rf_pred_smote_pca_class = predict(rf_smote_pca, newdata = test_data, type = "raw")
acc_rf_smote_pca = confusionMatrix(rf_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# Plot ROC curves
plot(roc_rf_no_smote_no_pca, col = "blue", main = "ROC for Random Forest")
lines(roc_rf_smote_no_pca, col = "red")
lines(roc_rf_no_smote_pca, col = "green")
lines(roc_rf_smote_pca, col = "purple")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 
       col = c("blue", "red", "green", "purple"), 
       lty = 1)
```

### Bagging

```{r}
#Bagging
# No SMOTE, No PCA
set.seed(123)
bag_no_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "treebag",
  trControl = ctrl_base, 
  metric = "ROC"
)
bag_pred_no_smote_no_pca = predict(bag_no_smote_no_pca, newdata = test_data, type = "prob")
roc_bag_no_smote_no_pca = roc(test_data$stroke, bag_pred_no_smote_no_pca[,"YES"])
bag_pred_no_smote_no_pca_class = predict(bag_no_smote_no_pca, newdata = test_data, type = "raw")
acc_bag_no_smote_no_pca = confusionMatrix(bag_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, No PCA
set.seed(123)
bag_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "treebag",
  trControl = ctrl_smote, 
  metric = "ROC"
)
bag_pred_smote_no_pca = predict(bag_smote_no_pca, newdata = test_data, type = "prob")
roc_bag_smote_no_pca = roc(test_data$stroke, bag_pred_smote_no_pca[,"YES"])
bag_pred_smote_no_pca_class = predict(bag_smote_no_pca, newdata = test_data, type = "raw")
acc_bag_smote_no_pca = confusionMatrix(bag_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# No SMOTE, PCA
set.seed(123)
bag_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "treebag",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
bag_pred_no_smote_pca = predict(bag_no_smote_pca, newdata = test_data, type = "prob")
roc_bag_no_smote_pca = roc(test_data$stroke, bag_pred_no_smote_pca[,"YES"])
bag_pred_no_smote_pca_class = predict(bag_no_smote_pca, newdata = test_data, type = "raw")
acc_bag_no_smote_pca = confusionMatrix(bag_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, PCA
set.seed(123)
bag_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "treebag",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  metric = "ROC"
)
bag_pred_smote_pca = predict(bag_smote_pca, newdata = test_data, type = "prob")
roc_bag_smote_pca = roc(test_data$stroke, bag_pred_smote_pca[,"YES"])
bag_pred_smote_pca_class = predict(bag_smote_pca, newdata = test_data, type = "raw")
acc_bag_smote_pca = confusionMatrix(bag_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# Plot ROC curves
plot(roc_bag_no_smote_no_pca, col = "blue", main = "ROC for Bagging")
lines(roc_bag_smote_no_pca, col = "red")
lines(roc_bag_no_smote_pca, col = "green")
lines(roc_bag_smote_pca, col = "purple")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 
       col = c("blue", "red", "green", "purple"), 
       lty = 1)
```

### Boosting

```{r}
#Boosting
# No SMOTE, No PCA
set.seed(123)
boost_no_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "gbm",
  trControl = ctrl_base, 
  tuneLength = 3,  # Tune n.trees, interaction.depth, etc.
  metric = "ROC",
  verbose = FALSE
)
boost_pred_no_smote_no_pca = predict(boost_no_smote_no_pca, newdata = test_data, type = "prob")
roc_boost_no_smote_no_pca = roc(test_data$stroke, boost_pred_no_smote_no_pca[,"YES"])
boost_pred_no_smote_no_pca_class = predict(boost_no_smote_no_pca, newdata = test_data, type = "raw")
acc_boost_no_smote_no_pca = confusionMatrix(boost_pred_no_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, No PCA
set.seed(123)
boost_smote_no_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "gbm",
  trControl = ctrl_smote, 
  tuneLength = 3,
  metric = "ROC",
  verbose = FALSE
)
boost_pred_smote_no_pca = predict(boost_smote_no_pca, newdata = test_data, type = "prob")
roc_boost_smote_no_pca = roc(test_data$stroke, boost_pred_smote_no_pca[,"YES"])
boost_pred_smote_no_pca_class = predict(boost_smote_no_pca, newdata = test_data, type = "raw")
acc_boost_smote_no_pca = confusionMatrix(boost_pred_smote_no_pca_class, test_data$stroke)$overall["Accuracy"]

# No SMOTE, PCA
set.seed(123)
boost_no_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "gbm",
  trControl = ctrl_base, 
  preProcess = c("center", "scale", "pca"),
  tuneLength = 3,
  metric = "ROC",
  verbose = FALSE
)
boost_pred_no_smote_pca = predict(boost_no_smote_pca, newdata = test_data, type = "prob")
roc_boost_no_smote_pca = roc(test_data$stroke, boost_pred_no_smote_pca[,"YES"])
boost_pred_no_smote_pca_class = predict(boost_no_smote_pca, newdata = test_data, type = "raw")
acc_boost_no_smote_pca = confusionMatrix(boost_pred_no_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# SMOTE, PCA
set.seed(123)
boost_smote_pca = train(
  stroke ~ ., 
  data = train_data, 
  method = "gbm",
  trControl = ctrl_smote, 
  preProcess = c("center", "scale", "pca"),
  tuneLength = 3,
  metric = "ROC",
  verbose = FALSE
)
boost_pred_smote_pca = predict(boost_smote_pca, newdata = test_data, type = "prob")
roc_boost_smote_pca = roc(test_data$stroke, boost_pred_smote_pca[,"YES"])
boost_pred_smote_pca_class = predict(boost_smote_pca, newdata = test_data, type = "raw")
acc_boost_smote_pca = confusionMatrix(boost_pred_smote_pca_class, test_data$stroke)$overall["Accuracy"]

# Plot ROC curves
plot(roc_boost_no_smote_no_pca, col = "blue", main = "ROC for Boosting")
lines(roc_boost_smote_no_pca, col = "red")
lines(roc_boost_no_smote_pca, col = "green")
lines(roc_boost_smote_pca, col = "purple")
legend("bottomright", 
       legend = c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 
       col = c("blue", "red", "green", "purple"), 
       lty = 1)
```
# Results

Every model-scenario pair is scored on the test set: AUC from the ROC curve, plus accuracy, sensitivity, and specificity from the confusion matrix. The table collects all four metrics; the bar plots after it make the trade-offs visual. QDA rows with failed fits are dropped.

```{r}
sens_logit_no_smote_no_pca = confusionMatrix(logit_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_logit_smote_no_pca = confusionMatrix(logit_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_logit_no_smote_pca = confusionMatrix(logit_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_logit_smote_pca = confusionMatrix(logit_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_lda_no_smote_no_pca = confusionMatrix(lda_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_lda_smote_no_pca = confusionMatrix(lda_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_lda_no_smote_pca = confusionMatrix(lda_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_lda_smote_pca = confusionMatrix(lda_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_qda_no_smote_no_pca = if (!is.null(qda_pred_no_smote_no_pca_class)) {
  confusionMatrix(qda_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
} else {
  NA
}
sens_qda_smote_no_pca = if (!is.null(qda_pred_smote_no_pca_class)) {
  confusionMatrix(qda_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
} else {
  NA
}
sens_qda_no_smote_pca = confusionMatrix(qda_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_qda_smote_pca = confusionMatrix(qda_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_dt_no_smote_no_pca = confusionMatrix(dt_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_dt_smote_no_pca = confusionMatrix(dt_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_dt_no_smote_pca = confusionMatrix(dt_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_dt_smote_pca = confusionMatrix(dt_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_rf_no_smote_no_pca = confusionMatrix(rf_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_rf_smote_no_pca = confusionMatrix(rf_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_rf_no_smote_pca = confusionMatrix(rf_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_rf_smote_pca = confusionMatrix(rf_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_bag_no_smote_no_pca = confusionMatrix(bag_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_bag_smote_no_pca = confusionMatrix(bag_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_bag_no_smote_pca = confusionMatrix(bag_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_bag_smote_pca = confusionMatrix(bag_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_boost_no_smote_no_pca = confusionMatrix(boost_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_boost_smote_no_pca = confusionMatrix(boost_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_boost_no_smote_pca = confusionMatrix(boost_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_boost_smote_pca = confusionMatrix(boost_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_knn_no_smote_no_pca = confusionMatrix(knn_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_knn_smote_no_pca = confusionMatrix(knn_pred_smote_no_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_knn_no_smote_pca = confusionMatrix(knn_pred_no_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
sens_knn_smote_pca = confusionMatrix(knn_pred_smote_pca_class, test_data$stroke)$byClass["Sensitivity"]
```
```{r}
spec_logit_no_smote_no_pca = confusionMatrix(logit_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_logit_smote_no_pca = confusionMatrix(logit_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_logit_no_smote_pca = confusionMatrix(logit_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_logit_smote_pca = confusionMatrix(logit_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_lda_no_smote_no_pca = confusionMatrix(lda_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_lda_smote_no_pca = confusionMatrix(lda_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_lda_no_smote_pca = confusionMatrix(lda_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_lda_smote_pca = confusionMatrix(lda_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_qda_no_smote_no_pca = if (!is.null(qda_pred_no_smote_no_pca_class)) {
  confusionMatrix(qda_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
} else {
  NA
}
spec_qda_smote_no_pca = if (!is.null(qda_pred_smote_no_pca_class)) {
  confusionMatrix(qda_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
} else {
  NA
}
spec_qda_no_smote_pca = confusionMatrix(qda_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_qda_smote_pca = confusionMatrix(qda_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_dt_no_smote_no_pca = confusionMatrix(dt_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_dt_smote_no_pca = confusionMatrix(dt_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_dt_no_smote_pca = confusionMatrix(dt_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_dt_smote_pca = confusionMatrix(dt_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_rf_no_smote_no_pca = confusionMatrix(rf_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_rf_smote_no_pca = confusionMatrix(rf_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_rf_no_smote_pca = confusionMatrix(rf_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_rf_smote_pca = confusionMatrix(rf_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_bag_no_smote_no_pca = confusionMatrix(bag_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_bag_smote_no_pca = confusionMatrix(bag_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_bag_no_smote_pca = confusionMatrix(bag_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_bag_smote_pca = confusionMatrix(bag_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_boost_no_smote_no_pca = confusionMatrix(boost_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_boost_smote_no_pca = confusionMatrix(boost_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_boost_no_smote_pca = confusionMatrix(boost_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_boost_smote_pca = confusionMatrix(boost_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_knn_no_smote_no_pca = confusionMatrix(knn_pred_no_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_knn_smote_no_pca = confusionMatrix(knn_pred_smote_no_pca_class, test_data$stroke)$byClass["Specificity"]
spec_knn_no_smote_pca = confusionMatrix(knn_pred_no_smote_pca_class, test_data$stroke)$byClass["Specificity"]
spec_knn_smote_pca = confusionMatrix(knn_pred_smote_pca_class, test_data$stroke)$byClass["Specificity"]
```
```{r}
results_df = data.frame(
  Model = rep(c("Logistic Regression", "LDA", "QDA", "Decision Tree", 
                "Random Forest", "Bagging", "Boosting", "KNN"), each = 4),
  Scenario = rep(c("No SMOTE No PCA", "SMOTE No PCA", "No SMOTE PCA", "SMOTE PCA"), 8),
  AUC = c(
    auc(roc_logit_no_smote_no_pca), auc(roc_logit_smote_no_pca), auc(roc_logit_no_smote_pca), auc(roc_logit_smote_pca),
    auc(roc_lda_no_smote_no_pca), auc(roc_lda_smote_no_pca), auc(roc_lda_no_smote_pca), auc(roc_lda_smote_pca),
    if (!is.null(roc_qda_no_smote_no_pca)) auc(roc_qda_no_smote_no_pca) else NA,
    if (!is.null(roc_qda_smote_no_pca)) auc(roc_qda_smote_no_pca) else NA,
    auc(roc_qda_no_smote_pca), auc(roc_qda_smote_pca),
    auc(roc_dt_no_smote_no_pca), auc(roc_dt_smote_no_pca), auc(roc_dt_no_smote_pca), auc(roc_dt_smote_pca),
    auc(roc_rf_no_smote_no_pca), auc(roc_rf_smote_no_pca), auc(roc_rf_no_smote_pca), auc(roc_rf_smote_pca),
    auc(roc_bag_no_smote_no_pca), auc(roc_bag_smote_no_pca), auc(roc_bag_no_smote_pca), auc(roc_bag_smote_pca),
    auc(roc_boost_no_smote_no_pca), auc(roc_boost_smote_no_pca), auc(roc_boost_no_smote_pca), auc(roc_boost_smote_pca),
    auc(roc_knn_no_smote_no_pca), auc(roc_knn_smote_no_pca), auc(roc_knn_no_smote_pca), auc(roc_knn_smote_pca)
  ),
  Accuracy = c(
    acc_logit_no_smote_no_pca, acc_logit_smote_no_pca, acc_logit_no_smote_pca, acc_logit_smote_pca,
    acc_lda_no_smote_no_pca, acc_lda_smote_no_pca, acc_lda_no_smote_pca, acc_lda_smote_pca,
    acc_qda_no_smote_no_pca, acc_qda_smote_no_pca, acc_qda_no_smote_pca, acc_qda_smote_pca,
    acc_dt_no_smote_no_pca, acc_dt_smote_no_pca, acc_dt_no_smote_pca, acc_dt_smote_pca,
    acc_rf_no_smote_no_pca, acc_rf_smote_no_pca, acc_rf_no_smote_pca, acc_rf_smote_pca,
    acc_bag_no_smote_no_pca, acc_bag_smote_no_pca, acc_bag_no_smote_pca, acc_bag_smote_pca,
    acc_boost_no_smote_no_pca, acc_boost_smote_no_pca, acc_boost_no_smote_pca, acc_boost_smote_pca,
    acc_knn_no_smote_no_pca, acc_knn_smote_no_pca, acc_knn_no_smote_pca, acc_knn_smote_pca
  ),
  Sensitivity = c(
    sens_logit_no_smote_no_pca, sens_logit_smote_no_pca, sens_logit_no_smote_pca, sens_logit_smote_pca,
    sens_lda_no_smote_no_pca, sens_lda_smote_no_pca, sens_lda_no_smote_pca, sens_lda_smote_pca,
    sens_qda_no_smote_no_pca, sens_qda_smote_no_pca, sens_qda_no_smote_pca, sens_qda_smote_pca,
    sens_dt_no_smote_no_pca, sens_dt_smote_no_pca, sens_dt_no_smote_pca, sens_dt_smote_pca,
    sens_rf_no_smote_no_pca, sens_rf_smote_no_pca, sens_rf_no_smote_pca, sens_rf_smote_pca,
    sens_bag_no_smote_no_pca, sens_bag_smote_no_pca, sens_bag_no_smote_pca, sens_bag_smote_pca,
    sens_boost_no_smote_no_pca, sens_boost_smote_no_pca, sens_boost_no_smote_pca, sens_boost_smote_pca,
    sens_knn_no_smote_no_pca, sens_knn_smote_no_pca, sens_knn_no_smote_pca, sens_knn_smote_pca
  ),
  Specificity = c(
    spec_logit_no_smote_no_pca, spec_logit_smote_no_pca, spec_logit_no_smote_pca, spec_logit_smote_pca,
    spec_lda_no_smote_no_pca, spec_lda_smote_no_pca, spec_lda_no_smote_pca, spec_lda_smote_pca,
    spec_qda_no_smote_no_pca, spec_qda_smote_no_pca, spec_qda_no_smote_pca, spec_qda_smote_pca,
    spec_dt_no_smote_no_pca, spec_dt_smote_no_pca, spec_dt_no_smote_pca, spec_dt_smote_pca,
    spec_rf_no_smote_no_pca, spec_rf_smote_no_pca, spec_rf_no_smote_pca, spec_rf_smote_pca,
    spec_bag_no_smote_no_pca, spec_bag_smote_no_pca, spec_bag_no_smote_pca, spec_bag_smote_pca,
    spec_boost_no_smote_no_pca, spec_boost_smote_no_pca, spec_boost_no_smote_pca, spec_boost_smote_pca,
    spec_knn_no_smote_no_pca, spec_knn_smote_no_pca, spec_knn_no_smote_pca, spec_knn_smote_pca
  )
) %>%
  filter(!is.na(AUC) & !is.na(Accuracy) & !is.na(Specificity))

kable(results_df, caption = "AUC, Accuracy,Sensitivity, and Specificity for Each Model and Scenario (NA Removed)", digits = 4)
```


```{r}
results_df$Scenario = factor(results_df$Scenario, 
                             levels = c("No SMOTE No PCA", "SMOTE No PCA", 
                                        "No SMOTE PCA", "SMOTE PCA"))

ggplot(results_df, aes(x = Model, y = Sensitivity, fill = Scenario)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.45)) +
  labs(title = "Sensitivity Comparison Across Models and Scenarios (NA Removed)",
       x = "Model",
       y = "Sensitivity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")
```

```{r}
results_df$Scenario = factor(results_df$Scenario, 
                             levels = c("No SMOTE No PCA", "SMOTE No PCA", 
                                        "No SMOTE PCA", "SMOTE PCA"))

ggplot(results_df, aes(x = Model, y = Specificity, fill = Scenario)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.45)) +
  labs(title = "Specificity Comparison Across Models and Scenarios (NA Removed)",
       x = "Model",
       y = "Specificity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")
```

Because the QDA model fails to fit in every fold due to the rank deficiency issue, no predictions are produced. Without predictions, the performance metrics (like ROC) cannot be computed, leading to the "missing values" warning.

```{r}
# Convert Scenario to factor for consistent ordering
results_df$Scenario = factor(results_df$Scenario, 
                         levels = c("No SMOTE No PCA", "SMOTE No PCA", 
                                    "No SMOTE PCA", "SMOTE PCA"))

# Create grouped bar plot
ggplot(results_df, aes(x = Model, y = AUC, fill = Scenario)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  labs(title = "AUC Comparison Across Models and Scenarios (NA Removed)",
       x = "Model",
       y = "AUC") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")
```


```{r}
# Accuracy Plot
ggplot(results_df, aes(x = Model, y = Accuracy, fill = Scenario)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.45)) +
  labs(title = "Accuracy Comparison Across Models and Scenarios (NA Removed)",
       x = "Model",
       y = "Accuracy") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set1")
```

# Takeaway

With only about 5% strokes in the data, accuracy alone misleads, which is why every model is judged on AUC, sensitivity, and specificity together. The four scenarios exist to price the two big choices: SMOTE rebalances the classes, PCA compresses the features, and the table shows what each costs per model. QDA could not fit in some folds because of rank deficiency, so those rows are out of the comparison. Read the table per metric, because the ranking moves with the metric, which is the honest answer for a rare-event problem like this.
